Kleene Categories

### 2025-07-05 kleenecat/kleenecat.tex
Different axiomatization of Kleene categories using parameterized list objects

\newcommand{\Set}{\cat{Set}}
\newcommand{\lto}{\multimap}
\newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}

@@

\section{Kleene Category}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axoims depicted
in Figure~\ref{fig:axioms}.

@@

The addition operation can be used to define the partial order
structure $a \leq b$ if $a + b = b$. In the theory of formal languages
this order structure can be used to model language containment. In this
section, we want to categorify the concept of Kleene algebra and
build on top of it in order to define an abstract theory of parsing.
We start by defining \emph{Kleene categories}.

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$
  such that for every objects $A$ and $B$, the endofunctors $F_{A, B}
  = B + A \otimes X$ and $G_{A, B} = B + X \otimes A$ have initial
  algebras (denoted $\mu X.\, F_{A, B}(X)$) such that the canonical
  morphisms $\mu X.\, F_{A, B}(X) \to B \otimes (\mu   X.\, F_{A, 1})$
  $\mu X.\, G_{A, B}(X) \to (\mu   X.\, G_{A, 1})\otimes B $ are iso.
\end{definition}

@@

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
\end{example}

An unexpected example comes from the theory of substructural logics.

@@

This proposed definition is sensible. After all, the proposed axioms
are a direct translation of the Kleene algebra axioms to a categorical
setting. Its most convoluted aspect is the axiomatization of the Kleene
star as a family os initial algebras satisfying certain
isomorphisms. If the Kleene category $\cat{K}$ has more structure,
then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    1}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : B
  \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
  by showing that the diagram on the left commute if, and only if,
  the diagram on the right commutes:

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds.
\end{proof}

@@

\begin{definition}
  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
  two functors $\cat{C} \to \Set$. Their Day convolution tensor
  product is defined as the following coend formula:
  \[
  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
  \]
  Dually, its internal hom is given by the following end formula:
  \[
  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
  \]
\end{definition}

@@

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}

  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts, and since the tensor
  is a left adjoint, it preserves colimits and, therefore, it is
  a distributive category.

@@

\begin{theorem}
  The morphism $\alpha : \grammar \to F(\grammar)$ is the final
  $F$-coalgebra.
\end{theorem}
\begin{proof}
  Given an $F$-coalgebra $\langle \alpha, \delta\rangle : \cat{C} \to \Set \times \cat{C}^\Sigma$, we can define the functor $\cat{C} \times \Sigma^* \to \Set$ much like in the DFA case:
  by induction on $\Sigma^*$.

\end{proof}

@@

For now, we ignore the subtleties behind pseudomonads, 2-monads,
lax algebras, etc. We begin by defining pre Kleene categories.

\begin{definition}
  A pre Kleene category is a distributive monoidal category such that
  the functors $F_{A, B}$ and $G_{A, B}$ have initial algebras, but
  they do not necessarily satisfy the isomorphism condition.
\end{definition}

These objects can pretty naturally be organized as a 2-category
$\cat{PKC}$: 1-cells are functors that strictly preserves all of the
structure while 2-cells are just natural transformations.

Let $\cat{C}$ be a category. It is more or less straightforward to
define free pre Kleene category $PKC(\cat{C})$. We can do that by
using the internal language of pre Kleene categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
inductive types corresponding to the initial algebras for $F_{A, B}$
and $G_{A, B}$, for every $A$, $B$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
We can prove directly by the equational theory of this type theory
that this is an inverse to the canonical map going in the other
direction. The inductive nature of type theories should make the
following theorem pretty straightforward to prove:
%
\begin{theorem}
  The forgetful functor $\cat{PKC} \to \cat{Cat}$ has a left adjoint
  in $\cat{2Cat}$.
\end{theorem}

Unfortunately, we can see that such a free construction will not be a
Kleene category, since there is no way of constructing the Kleene star
isomorphisms. That being said, there is a comparison map
$\mu F_{A,B} \to A \mu F_{1, B}$ for $F$ and $G$. This leads me to conjecture
the following:
%
\begin{conjecture}
  The forgetful functor $\cat{KC} \to \cat{PKC}$ is a reflective
  subcategory
\end{conjecture}

If the theorem above is true, we get a Kleene category
monad on $\cat{Cat}$ by composing the adjunctions above.
While I'm not sure how to define the reflection, Theorem~\ref{th:kleeneclosed}
might be important in our construction. I remember asking Daniel Gratzer
about this and he told me something about mapping a pre Kleene
category $\cat{C}$ to $\widehat{\cat{C}}$ and ``carving it out'' in order
to obtain an appropriate subcategory. Note that $\widehat{C}$ is monoidal
closed, so it will be Kleene as well.

>>>>

\newcommand{\Set}{\cat{Set}}
\newcommand{\lto}{\multimap}
\newcommand{\grammar}{\Set^{\cat{\Sigma^*}}}
\newtheorem*{construction}{Construction}

@@

\section{Kleene Algebra and Kleene Categories}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, a Kleene algebra is a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axoims depicted
in Figure~\ref{fig:axioms}.

@@

The addition operation induces a partial order structure where $a \leq
b$ if $a + b = b$. In the theory of formal languages this order
structure models language containment.

\begin{example}
  The free Kleene algebra over a set $\Sigma$ is given by regular
  expressions with alphabet $\Sigma$, quotiented by the Kleene algebra
  axioms.
\end{example}

Kleene algebra provides an algebraic foundation for the implementation
of langauge \emph{recognizers}, i.e., programs that determine whether
or not a string is a member of a formal language. On the other hand,
the fact that $+$ is idempotent means that a Kleene algebra cannot
distinguish between multiple parses of the same string, and so Kleene
algebra is insufficient for the discussion of \emph{parsers} which may
select one of several different possible parses in the case of
ambiguous grammars.

Our goal is to provide a generalization of Kleene algebra that is
``proof-relevant'', and so can provide a foundation for parsing
regular expressions in the same way that Kleene algebra does for
recognition. To do this we need to move beyond the view of a Kleene
algebra as algebraic structure over sets, and instead as a kind of
structured \emph{poset}. We then generalize this to a Kleene
\emph{category}, where the morphisms of the category correspond to not
just language inclusion but \emph{formal reductions} of grammars, and
isomorphism is not language equivalence but instead \emph{strong
equivalence} of grammars.

The most complex part of this generalization is the appropriate
generalization of the Kleene star. For that we make the following
definitions
\begin{definition}
  Let $\cat C$ be a monoidal category. Given an object $A \in \cat C$,
  a \textbf{parameterized cons-list object} consists of
  \begin{enumerate}
  \item an object $L \in \cat C$
  \item with morphisms $\textrm{nil} : I \to L$ and $\textrm{cons} : A \otimes L \to L$
  \item such that for every $P,B \in \cat C$, and morphisms $n : P \to
    B$ and $c : A \otimes B \to B$ there is a unique morphism $\textrm{foldr}(n,c) : L
    \otimes P \to B$ satisfying

    % https://q.uiver.app/#q=WzAsNixbMSwwLCJMIFxcb3RpbWVzIFAiXSxbMCwwLCJJIFxcb3RpbWVzIFAiXSxbMCwxLCJQIl0sWzIsMCwiQSBcXG90aW1lcyBMIFxcb3RpbWVzIFAiXSxbMiwxLCJBIFxcb3RpbWVzIEIiXSxbMSwxLCJCIl0sWzEsMiwiIiwwLHsibGV2ZWwiOjIsInN0eWxlIjp7ImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMiw1LCJuIl0sWzQsNSwiYyIsMl0sWzMsNCwiQSBcXG90aW1lcyBcXHRleHRybXtmb2xkcn0obixjKSJdLFszLDAsIlxcdGV4dHJte2NvbnN9IFxcb3RpbWVzIFAiLDJdLFsxLDAsIlxcdGV4dHJte25pbH0gXFxvdGltZXMgUCJdLFswLDUsIlxcdGV4dHJte2ZvbGRyfShuLGMpIl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{I \otimes P} \& {L \otimes P} \& {A \otimes L \otimes P} \\
	P \& B \& {A \otimes B}
	\arrow["{\textrm{nil} \otimes P}", from=1-1, to=1-2]
	\arrow[equals, from=1-1, to=2-1]
	\arrow["{\textrm{foldr}(n,c)}", from=1-2, to=2-2]
	\arrow["{\textrm{cons} \otimes P}"', from=1-3, to=1-2]
	\arrow["{A \otimes \textrm{foldr}(n,c)}", from=1-3, to=2-3]
	\arrow["n", from=2-1, to=2-2]
	\arrow["c"', from=2-3, to=2-2]
\end{tikzcd}\]
  \end{enumerate}

  A \textbf{cons-list object} is like a parameterized cons-list object but where $P$ can only be $I$.

  A \textbf{parameterized snoc-list object} is a cons-list object in
  the dual monoidal structure $\cat C^{co}$. In this case we call the
  $\textrm{cons}$ instead $\textrm{snoc}$ and the $\textrm{foldr}$
  morphism $\textrm{foldl}$. A non-parameterized snoc-list object is
  defined similarly.
\end{definition}

\begin{construction}
  Given a cons-list object for $A$, we can define a morphism $\textrm{snoc} : L \otimes A \to L$ as $\textrm{foldr}(\eta, \textrm{cons})$
  where $\eta : A \to L$ is the composition
  % https://q.uiver.app/#q=WzAsNCxbMCwwLCJBIl0sWzEsMCwiQSBcXG90aW1lcyBJIl0sWzIsMCwiQSBcXG90aW1lcyBMIl0sWzMsMCwiTCJdLFswLDEsIiIsMCx7ImxldmVsIjoyLCJzdHlsZSI6eyJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMiwiQSBcXG90aW1lcyBcXHRleHRybXtuaWx9Il0sWzIsMywiXFx0ZXh0cm17Y29uc30iXV0=
\[\begin{tikzcd}[ampersand replacement=\&]
	A \& {A \otimes I} \& {A \otimes L} \& L
	\arrow[equals, from=1-1, to=1-2]
	\arrow["{A \otimes \textrm{nil}}", from=1-2, to=1-3]
	\arrow["{\textrm{cons}}", from=1-3, to=1-4]
\end{tikzcd}\]
\end{construction}

\begin{definition}
  A Kleene category is a distributive monoidal category $\cat{K}$ such
  that for every object $A$, there is a cons-list object
  $(A^*,\textrm{nil},\textrm{cons})$ such that $(A^*, \textrm{nil},
  \textrm{snoc})$ is a snoc-list object.
\end{definition}

@@

\begin{lemma}
  A Kleene algebra is equivalent to a posetal Kleene category.
\end{lemma}

% This isn't an example of a Kleene category, it's an example of a model of conjunctive ordered logic
An unexpected example comes from the theory of substructural logics.

@@

This proposed definition is sensible. After all, the proposed axioms
are a direct translation of the Kleene algebra axioms to a categorical
setting. Its most convoluted aspect is the axiomatization of the
Kleene star as a parameterized list and snoc object. If the Kleene
category $\cat{K}$ has more structure, then these isomorphisms hold
``for free''.

\begin{definition}
  A pre-Kleene category is a distributive monoidal category with a (non-parameterized) cons-list object.
\end{definition}

\begin{theorem}
  If $\cat K$ is a pre-Kleene category that is also biclosed, then it is a Kleene category.
\end{theorem}
\begin{proof}
  TODO: closed implies parameterized for free and that cons-list and snoc-list objects are equivalent.
\end{proof}

%% \begin{theorem}
%%   \label{th:kleeneclosed}
%%   Let $\cat{K}$ be a Kleene category such that it is also monoidal
%%   closed.  Then, the initial algebras isomorphisms hold automatically.
%% \end{theorem}
%% \begin{proof}
%%   We prove this by the unicity (up-to isomorphism) of initial
%%   algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
%%   (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
%%   X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
%%   (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
%%     1}(X))$.

%%   Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
%%   we want to show that there is a unique algebra morphism $h : B
%%   \otimes \mu X.\, F_{A,1} \to Y$. We can show existence and uniqueness
%%   by showing that the diagram on the left commute if, and only if,
%%   the diagram on the right commutes:

%%   This equivalence follows by using the adjunction structure given
%%   by the monoidal closed structure of $\cat{K}$. A completely analogous
%%   argument for $G_{A,B}$ also holds.
%% \end{proof}

@@

\begin{definition}
  Let $\cat{C}$ be a locally small monoidal category and $F$, $G$ be
  two functors $\cat{C} \to \Set$. Their Day convolution tensor
  product is defined as the following coend formula:
  \[
  (F \otimes_{Day} G)(x) = \int^{(y,z) \in \cat{C}\times\cat{C}}\cat{C}(y\otimes z, x) \times F(y) \times G(z) 
  \]
  Dually, its internal hom is given by the following end formula:
  \[
  (F \lto_{Day} G)(x) = \int_{y} \Set(F(y), G(x \otimes y))
  \]
\end{definition}
Note that in the case the category $\cat C$ is the discrete category
of a monoid, the Day convolution admits a simpler description where
the coend degenerates to a coproduct and the end ot a product.

@@

%% \begin{theorem}
%%   Let $\cat{K}$ be a Kleene category and $A$ a discrete category.
%%   The functor category $[A, \cat{K}]$.
%%   (HOW GENERAL SHOULD THIS THEOREM BE? BY ASSUMING ENOUGH STRUCTURE,
%%   E.G. K = Set, THIS THEOREM BECOMES SIMPLE TO PROVE)
%% \end{theorem}
\begin{theorem}
  If $\cat{C}$ is a locally small monoidal category, then
  $\Set^{\cat{C}}$ is a Kleene category.
\end{theorem}
\begin{proof}
  By the lemma above, $\Set^{\cat{C}}$ is monoidal closed, and since it
  is a presheaf category, it has coproducts, and since the tensor
  is a left adjoint, it preserves colimits and, therefore, it is
  a distributive category.

@@

\begin{theorem}
  The morphism $\alpha : \grammar \to F(\grammar)$ is the final
  $F$-coalgebra.
\end{theorem}
\begin{proof}
  Given an $F$-coalgebra $\langle \alpha, \delta\rangle : \cat{C} \to \Set \times \cat{C}^\Sigma$, we can define the functor $\cat{C} \times \Sigma^* \to \Set$ much like in the DFA case:
  by induction on $\Sigma^*$.
\end{proof}

@@

For now, we ignore the subtleties behind pseudomonads, 2-monads,
lax algebras, etc.

Let $\cat{C}$ be a category. It is more or less straightforward to
define the free Kleene category $PKC(\cat{C})$. We can do that by
using the internal language of Kleene categories. Concretely, it
will be a type theory with (non-commutative) tensors, sum types and
a primitive type of lists $L(A)$ for every $A$. By type theoretic reasons, this
structure forces the coproduct to distribute over the tensor, where we
can write the morphism $A \otimes (B + C) \to (A \otimes B) + (A
\otimes C) $ as the term
\begin{align*}
  &z \vdash \mathsf{let}\, x \otimes y = z\, \mathsf{in}\\
  &\mathsf{case}\, y \, \mathsf{with}\\
  &| \mathsf{in}_1\, y_1 \Rightarrow \mathsf{in}_1\, (x \otimes y_1)\\
  &| \mathsf{in}_2\,y_2\Rightarrow \mathsf{in}_2 \,(x \otimes y_2)\\
  &\mathsf{end}
\end{align*}
%
The list type includes primitive terms for $\textrm{cons}$,
$\textrm{nil}$, $\textrm{foldr}$ and $\textrm{foldl}$ where
$\textrm{foldr}$ and $\textrm{foldl}$ allow for a context to be used
in the construction of the nil case, but no free variables can be used
in the cons case:
\begin{mathpar}
  \inferrule
  {\Gamma \vdash L : A^* \and \Delta \vdash N : B \and a:A,b:B \vdash M : B
  }
  {\Gamma,\Delta \vdash \textrm{foldr}(L,N,a.b.M) : B}
\end{mathpar}

We can prove directly by the equational theory of this type theory
that this is an inverse to the canonical map going in the other
direction. The inductive nature of type theories should make the
following theorem pretty straightforward to prove:
%
\begin{theorem}
  The forgetful functor $\cat{KC} \to \cat{Cat}$ has a left adjoint
  in $\cat{2Cat}$.
\end{theorem}
